Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-06-20 paper.tex
final grammar/typography check

\section{Introduction}
Scan is one of the most ubiquitous operations in GPU computing, with applications in sorting~\cite{adinets2022onesweepfastersignificantdigit}, sparse matrix operations~\cite{BellGarland2009}, and more. The current state-of-the-art prefix scan algorithm, Merrill and Garland's \emph{Decoupled Lookback}~\cite{Merrill2016}, achieves near-ideal performance on NVIDIA hardware via CUDA by leveraging the \emph{Chained Scan} architecture. This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$. \emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{speed-of-light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device. However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward-progress guarantee}; without this guarantee, it risks indefinite \emph{starvation} and will never complete. In practice, it is easy to demonstrate this highly undesirable behavior.

@@

  \caption{The four global scan techniques described in \ref{sec:Evolution}. Dashed grey boxes indicate individual kernel launches.}
  \label{fig:four-techniques}
\end{figure*}

@@

Hardware-level considerations underscore the limitations of the fixed-launch approach. On GPUs where the global memory data cache is indexed by physical addresses—and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)—fixed launches can lead to performance degradation due to cache thrashing. This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups. Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns—or applications like radix sorting, where the scattering phase of each pass results in $radixDigits \cdot o$ unique memory locations—may experience significant performance degradation.

@@

\begin{figure*}[t]
  \centering
  \includegraphics[width=.95\textwidth]{graphics/DecoupledFallback.pdf}
  \caption{\emph{Decoupled Fallback}. For brevity, the chained propagation phase is condensed, see \ref{fig:decoupled-lookback} for the full behavior.}
  \label{fig:decoupled-fallback}
\end{figure*}

@@

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$. We find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4. Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@. For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum. Our test hardware, shown in Figure~\ref{tab:hardware}, spans a wide range of FPG and performance capabilities. We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing.
\begin{table*}
  \centering
  \begin{tabular}{l l c r r}
    \toprule
    Vendor & Device Name            & FPG         & Occupancy & Subgroup Size \\
    \midrule
    Intel  & HD Graphics 620        & \textbf{Yes} & 9         & 8--32         \\
    ARM    & Mali-G78 MP20          & \textbf{No}  & 40        & 16            \\
    Apple  & M3     (10 Cores)      & \textbf{No}  & 80        & 32            \\
    Apple  & M1 Max (32 Cores)      & \textbf{No}  & 160       & 32            \\
    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192       & 32            \\
    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420       & 32--64        \\
    \bottomrule
  \end{tabular}
  \caption{Hardware Specifications of Testing Devices; FPG determinations are based on the data collected by Sorensen et. al.~\cite{sorensen2021}\label{tab:hardware}}
\end{table*}

@@

As shown in Table~\ref{tab:results}, \emph{DF} achieves at least a $\sim$40\% speedup over \emph{RTS} across all tested devices, closely approaching the theoretically expected $O(3n)/O(2n)$ 50\% speedup, except on the 7900XT and Mali-G78. This result initially appears confusing, given that \emph{DF} outperforms \emph{Memcpy} by 6\% and 13\%, respectively. However, we hypothesized and confirmed a secondary effect of \emph{Chained Scans}: they enforce a queue-like behavior among workgroups. In-flight workgroups with a lower relative rank traverse fewer tiles during their lookbacks, enabling them to exit earlier than higher-ranked workgroups. Because this ordering emerges naturally from differing amounts of computation rather than explicit synchronization, the exit sequence is \emph{k-sorted} or \emph{nearly ordered}, rather than strictly ordered. Scheduling fairness and message-passing visibility can further influence this sequence---for example, an unfair scheduler might allocate more resources to a higher-ranked workgroup, causing it to exit before a lower-ranked counterpart, or a higher-ranked workgroup might observe an updated flag sooner, enabling earlier completion. We confirmed this behavior experimentally by using atomics to record and print the order in which workgroups completed their lookback phase. Unlike a simple \emph{Memcpy} kernel, where all in-flight workgroups begin writing immediately after completing reads, this “ordered exit” pattern facilitates more efficient global memory accesses during the write phase of the scan, as adjacent workgroups are more likely to share memory pages, reducing premature page evictions. We hypothesize that the 7900XT and Mali-G78 have insufficient memory systems to efficiently accommodate all in-flight workgroups, making them particularly sensitive to this effect. To test this hypothesis, we augmented our \emph{Memcpy} kernel with a simplified lookback routine---excluding the scan itself---and observed a performance improvement of approximately 10\% on the 7900XT. Thus, the unexpectedly slow baseline \emph{Memcpy} performance on these devices explains why \emph{DF}, despite its strong overall performance, does not fully achieve the theoretical 50\% speedup over \emph{RTS}.

@@

  \item \textbf{Total Fallbacks Initiated:} The total number of fallback operations initiated across all workgroups. Ideally, this number would be zero on devices with FPG; because multiple workgroups may perform a fallback on the same blocking tile, this number can exceed the number of workgroups launched. In practice, this count is highly dependent on the fairness of the hardware scheduler. On most devices tested, initiated fallbacks remain modest---about 10\% or fewer of the total workgroups launched.

@@

\subsection{Future Work}
Sorting is one of the most notable beneficiaries of single-pass scan and, by extension, \emph{Decoupled Fallback}. The current state-of-the-art algorithm, Adinets and Merrill's \emph{Onesweep}~\cite{adinets2022onesweepfastersignificantdigit}, is an 8-bit least-significant-digit radix sort, where single-pass scan minimizes data movement during digit-binning passes. However, due to its compute-intensive nature and large $radixDigit$-sized monoid, it is unclear whether \emph{Decoupled Fallback} will map well to sorting or if new techniques are needed.

@@

Our artifact is available in the following GitHub repository: https://github.com/b0nes164/Decoupled-Fallback-Paper. The repository contains per API requirements and building instructions.

\begin{acks}
We thank Elias Naur for his initial work on Scalar Fallback, Ken Russell for facilitating this colloboration, Duane Merrill for answering our questions about scan terminology, and Oliver Le for providing the initial testing on Apple devices.
\end{acks}

>>>>

\section{Introduction}
Scan is one of the most ubiquitous operations in GPU computing, with applications in sorting~\cite{adinets2022onesweepfastersignificantdigit}, sparse matrix operations~\cite{BellGarland2009}, and more. The current state-of-the-art prefix scan algorithm, Merrill and Garland's \emph{Decoupled Lookback}~\cite{Merrill2016}, achieves near-ideal performance on NVIDIA hardware via CUDA by leveraging the \emph{Chained Scan} architecture. This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$.\ \emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{speed-of-light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device. However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward-progress guarantee}; without this guarantee, it risks indefinite \emph{starvation} and will never complete. In practice, it is easy to demonstrate this highly undesirable behavior.

@@

  \caption{The four global scan techniques described in Section~\ref{sec:Evolution}. Dashed grey boxes indicate individual kernel launches.}
  \label{fig:four-techniques}
\end{figure*}

@@

Hardware-level considerations underscore the limitations of the fixed-launch approach. On GPUs where the global memory data cache is indexed by physical addresses—and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)—fixed launches can lead to performance degradation due to cache thrashing. This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups. Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns—or applications like radix sorting, where the scattering phase of each pass results in $\textit{radixDigits} \cdot o$ unique memory locations—may experience significant performance degradation.

@@

\begin{figure*}[t]
  \centering
  \includegraphics[width=.95\textwidth]{graphics/DecoupledFallback.pdf}
  \caption{\emph{Decoupled Fallback}. For brevity, the chained propagation phase is condensed; see Figure~\ref{fig:decoupled-lookback} for the full behavior.}
  \label{fig:decoupled-fallback}
\end{figure*}

@@

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$. We find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4. Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@. For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum. Our test hardware, shown in Figure~\ref{tab:hardware}, spans a wide range of FPG and performance capabilities. We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing.
\begin{table*}
  \centering
  \begin{tabular}{l l c r r}
    \toprule
    Vendor & Device Name            & FPG         & Occupancy & Subgroup Size \\
    \midrule
    Intel  & HD Graphics 620        & \textbf{Yes} & 9         & 8--32         \\
    ARM    & Mali-G78 MP20          & \textbf{No}  & 40        & 16            \\
    Apple  & M3     (10 Cores)      & \textbf{No}  & 80        & 32            \\
    Apple  & M1 Max (32 Cores)      & \textbf{No}  & 160       & 32            \\
    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192       & 32            \\
    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420       & 32--64        \\
    \bottomrule
  \end{tabular}
  \caption{Hardware Specifications of Testing Devices; FPG determinations are based on the data collected by Sorensen et al.~\cite{sorensen2021}.\label{tab:hardware}}
\end{table*}

@@

As shown in Table~\ref{tab:results}, \emph{DF} achieves at least a $\sim$40\% speedup over \emph{RTS} across all tested devices, closely approaching the theoretically expected $O(3n)/O(2n)$ 50\% speedup, except on the 7900XT and Mali-G78. This result initially appears confusing, given that \emph{DF} outperforms \emph{Memcpy} by 6\% and 13\%, respectively. However, we hypothesized and confirmed a secondary effect of \emph{Chained Scans}: they enforce a queue-like behavior among workgroups. In-flight workgroups with a lower relative rank traverse fewer tiles during their lookbacks, enabling them to exit earlier than higher-ranked workgroups. Because this ordering emerges naturally from differing amounts of computation rather than explicit synchronization, the exit sequence is \emph{k-sorted} or \emph{nearly ordered}, rather than strictly ordered. Scheduling fairness and message-passing visibility can further influence this sequence---for example, an unfair scheduler might allocate more resources to a higher-ranked workgroup, causing it to exit before a lower-ranked counterpart, or a higher-ranked workgroup might observe an updated flag sooner, enabling earlier completion. We confirmed this behavior experimentally by using atomics to record and print the order in which workgroups completed their lookback phase. Unlike a simple \emph{Memcpy} kernel, where all in-flight workgroups begin writing immediately after completing reads, this “ordered exit” pattern facilitates more efficient global memory accesses during the write phase of the scan, as adjacent workgroups are more likely to share memory pages, reducing premature page evictions. We hypothesize that the 7900XT and Mali-G78 have insufficient memory systems to efficiently accommodate all in-flight workgroups, making them particularly sensitive to this effect. To test this hypothesis, we augmented our \emph{Memcpy} kernel with a simplified lookback routine---excluding the scan itself---and observed a performance improvement of approximately 10\% on the 7900XT\@. Thus, the unexpectedly slow baseline \emph{Memcpy} performance on these devices explains why \emph{DF}, despite its strong overall performance, does not fully achieve the theoretical 50\% speedup over \emph{RTS}.

@@

  \item \textbf{Total Fallbacks Initiated:} The total number of fallback operations initiated across all workgroups. Ideally, this number would be zero on devices with FPG\@; because multiple workgroups may perform a fallback on the same blocking tile, this number can exceed the number of workgroups launched. In practice, this count is highly dependent on the fairness of the hardware scheduler. On most devices tested, initiated fallbacks remain modest---about 10\% or fewer of the total workgroups launched.

@@

\subsection{Future Work}
Sorting is one of the most notable beneficiaries of single-pass scan and, by extension, \emph{Decoupled Fallback}. The current state-of-the-art algorithm, Adinets and Merrill's \emph{Onesweep}~\cite{adinets2022onesweepfastersignificantdigit}, is an 8-bit least-significant-digit radix sort, where single-pass scan minimizes data movement during digit-binning passes. However, due to its compute-intensive nature and large $\textit{radixDigit}$-sized monoid, it is unclear whether \emph{Decoupled Fallback} will map well to sorting or if new techniques are needed.

@@

Our artifact is available in the following GitHub repository: https://github.com/b0nes164/Decoupled-Fallback-Paper. The repository contains per-API requirements and building instructions.

\begin{acks}
We thank Elias Naur for his initial work on Scalar Fallback, Ken Russell for facilitating this collaboration, Duane Merrill for answering our questions about scan terminology, and Oliver Le for providing the initial testing on Apple devices.
\end{acks}
